Staged Abstract Interpreters

### 2019-02-12 paper/10_intro.tex
revise intro

Roughly at the same time when Futamura projections was visioned in 1970s,
\citeauthor{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fixpoints \cite{DBLP:conf/popl/CousotC77}.
Recent progress such as Abstracting Abstract Machines (AAM) uncovers a
methodology to derive sound abstract interpreters from their concrete
counterparts, and it has been applied to different variants of definitional
interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.
Given the structural similarity between concrete and abstract interpreters,
intellectually it is natural to raise the question whether it is possible to
specialize an abstract interpreter, and how can we effectively do that having
considered their different functionalities.

This paper studies the confluence of two old ideas --- Futamura projection and
abstract interpretation, but from the perspective of their recent realizations
--- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract interpreters.
By borrowing the ideas from \note{monadic interpreters} and embedded
domain-specific languages, we develop an approach to fulfill four different
semantics (Figure~\ref{confluence}) in a unified framework.

@@

\note{besides the intellectual merit of the confluence diagram, practical
problem: can we speed up a static analyzer without pervasive changes of the
abstract semantics, even without changing a single line of the abstract
interpreter?}

>>>>

\todo{example of interpreter -> compiler, mix}

Roughly at the same time when Futamura projections was visioned in 1970s,
\citeauthor{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fixpoints \cite{DBLP:conf/popl/CousotC77}. However,
constructing sound abstract interpreters was considered abstruse and complicated
for a long time.
Recent progress such as Abstracting Abstract Machines (AAM)
uncovers a systematic method to derive sound abstract interpreters from their
concrete counterparts, where the soundness can be easily established by
examining the transformation of semantics.
The AAM approach also has been been applied to different variants of
definitional interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.
\todo{CEK example?}

Now we have seen two orthogonal transformations of interpreters: by
speclization, we can refactor interpreters to compilers; by abstraction, we can
refactor interpreters to static analyzers. As an intellectual quest, it is
natural to ask, starting from an interpreter, can we derive sound abstract
interpreters but it works like a compiler and generates fast analysis code?

%artifacts of concrete semantics can be refactored to abstract interpreters.
In this paper, we study the confluence of two old ideas --- Futamura projection and
abstract interpretation, but from the perspective of their recent realizations
--- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract interpreters.
By borrowing the ideas from \note{monadic interpreters} and embedded
domain-specific languages, we develop an approach to fulfill four different
semantics (Figure~\ref{confluence}) in a unified framework. \todo{explain four sem}

@@

Besides the intellectual merit of the confluence diagram, this paper contributes
to a practical problem in static analysis: can we speed up a static analyzer
without any intrusive changes and soundness compromises?
